---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dpmsplit

<!-- badges: start -->
[![R-CMD-check](https://github.com/ONSdigital/dpmsplit/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ONSDigital/dpmsplit/actions/workflows/R-CMD-check.yaml)

[![Style](https://github.com/ONSdigital/dpmsplit/actions/workflows/style.yaml/badge.svg)](https://github.com/ONSDigital/dpmsplit/actions/workflows/style.yaml)

[![Contributor Covenant](https://img.shields.io/badge/Contributor%20Covenant-2.1-4baaaa.svg)](code_of_conduct.md)
<!-- badges: end -->

## Overview

Package **dpmsplit** is for taking estimates of total inflows or total outflows and splitting them out into component flows. For instance, **dpmsplit** might be used to turn estimates of total moves out of a region into (i) moves from the region to elsewhere in the country and (ii) moves from the region to overseas.


## Contributor guidance

**To see the complete [contributor guidance](https://github.com/ONSdigital/dpmsplit/blob/main/CONTRIBUTING.md)**

**In brief:**
All modifications to this package must be made through **Pull requests**, ideally linked to a specific GitHub **Issue**. 

Pull requests must be reviewed by someone other than the requester, if more than one individual was responsible for the requested changes the reviewer must be someone other than the main contributor, ideally it will be reviewed by someone entirely uninvolved with the particular changes made. 

Pull requests should have a suitable pull request review form attached, with key notes/descriptions of changes made added by the requester and if applicable additional notes on particular areas to focus the review (syntax changes, key functional changes, documentational changes etc.)


## Getting started

### Installation
#### Install directly from GitHub

If you have linked your RStudio installation to GitHub you should be able to install the package directly from GitHub using the install_github method from the devtools package

``` {r installation_github, eval=FALSE}
# install.packages("devtools")
library(devtools)

devtools::install_github("ONSdigital/dpmsplit", build_vignettes = TRUE, INSTALL_opts = "--no-multiarch")
```

If you have not/are unable to link your RStudio installation to GitHub (you may encounter a 404 error when attempting the previous approach) you can also install the package in two alternative ways

#### Install locally from .zip

Download a copy of the package repository as a .zip file (option in 'Code', below Open with GitHub Desktop) and install using the install_local method from the devtools package (replace the path with the path to the .zip file download location)

``` {r installation_zip, eval=FALSE}
# install.packages("devtools")
library(devtools)

devtools::install_local("C:/.../Downloads/dpmsplit-main.zip", build_vignettes = TRUE, INSTALL_opts = "--no-multiarch")
```

#### Install locally from cloned repository

Clone the repository (options in 'Code' to HTTPS/SSH paths) and build the package using the build() method from the devtools package

``` {r installation_clone, eval=FALSE}
# install.packages("devtools")
library(devtools)

dpmsplit_build <- devtools::build("~/put/the/package/path/here")

devtools::install_local(dpmsplit_build, build_vignettes = TRUE, INSTALL_opts = "--no-multiarch")
```

## Usage (example)

The two key functions are `split_single` and `split_multi`. `split_single` is for splitting for a single population,

```{r example}
library(dpmsplit)
res <- split_single(
  totals = sim_totals,
  reported = sim_reported,
  zeros_to_ones = TRUE
)
```

and `split_multi` is for splitting flows between multiple populations.
